Autopolyploid establishment through polygenic adaptation

### 2025-03-13 doc/paper/paper.tex
starting revisions

\documentclass[12pt,a4paper]{article}
\input{preamble.tex}
\usepackage[margin=1in]{geometry}
\usepackage{authblk}
\externaldocument[s-]{supp}

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics (\textit{sensu}
    \cite{barton2017}) for the inheritance of an additive quantitative trait in
    a mixed-ploidy population consisting of diploid, triploid and
    autotetraploid individuals producing haploid and diploid gametes.  We
    implement efficient simulation methods and use these to study the
    quantitative genetics of mixed-ploidy populations and the establishment of
    autotetraploids in a new habitat.  We show that for rare migration from the
    source population, autotetraploids are more likely to found a successful
    population than diploids under a very broad range of conditions, but that
    this is unlikely to sufficiently counter the scarcity of tetraploid
    founders when the source is predominantly diploid.  We assess in more
    detail how minority cytotype exclusion interacts with migration load in the
    establishment process and evaluate the impact of additional sources of
    prezygotic isolation, specifically selfing and assortative mating, on the
    relative establishment probabilities of the different cytotypes.
    In the discussion we consider how inbreeding depression may impact our
    findings.

@@

% Should connect more with literature and concepts of isolating mechanisms pre
% vs. postzygotic isolation, and reinforcement etc. (see e.g. Parvathy &
% Himani's paper introduction)

@@

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population through local extinction and 
recolonization dynamics has been suggested as a plausible model
\citep{lande1985}.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for local tetraploid establishment to
occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of \textit{prezygotic isolation} between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
localized dispersal \citep{baack2005,kolar2017}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression \citep{ronfort1999,otto2000}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

@@

Our notation is summarized in \cref{tbl:glossary}.
We consider a mixed-ploidy population of size $N$ consisting of $N_2$ diploid,
$N_3$ triploid and $N_4=N-N_2-N_3$ tetraploid individuals.
We assume an individual of ploidy level $k$ forms haploid and diploid gametes
with proportions $u_{k1}$ and $u_{k2}$, as well as a proportion
$1-u_{k1}-u_{k2}$ inviable (e.g. aneuploid or polyploid) gametes.
The (relative) fecundity of a $k$-ploid individual is hence $u_{k1} + u_{k2}$.
Unless stated otherwise, we will assume 
\begin{equation}
    \begin{pmatrix} 
    u_{21} & u_{22} \\ 
    u_{31} & u_{32} \\ 
    u_{41} & u_{42} 
    \end{pmatrix} =
    \begin{pmatrix} 
    1-u & u \\
    v & v \\
    0 & 1-u
    \end{pmatrix} \label{eq:U}
\end{equation}
where $u$ is referred to as the proportion of unreduced gametes, and $2v$ is
the proportion of euploid gametes produced by a triploid individual.

@@

If we would naively assume that the allelic effects underlying an additive
trait are identical across ploidy levels, 
a tetraploid offspring from a cross between two diploids would have, on average,
a trait value which is the sum of the parental trait values.
This is not likely to reflect biological reality: tetraploids do not tend to
have, for instance, twice the size of their diploid progenitors on average.
Furthermore, the genetic variance at Hardy-Weinberg and linkage equilibrium
(HWLE) in a large non-inbred tetraploid population will be twice that of their
diploid counterparts under such assumptions, which is similarly unrealistic
\citep{clo2022}.

@@

Again following \cite{barton2018}, we simulate the model by first calculating
the fitness of each parental pair $(i,j)$, which is the expected fitness of
offspring of this pair
\begin{equation}
  w_{ij}
    = \sum_{k=1}^2\sum_{l=1}^2 w_{ij}^{kl}
    = \sum_{k=1}^2\sum_{l=1}^2 u_{c_{i},k}u_{c_{j},l}
        \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right]
\end{equation}
The expectation on the right hand side can be calculated from \cref{eq:oneline}
using the moment-generating function of the Gaussian.
Having calculated the $w_{ij}$, the number of offspring surviving into the next
generation is calculated as $N^\ast(t+1) = \sum_{i,j}w_{ij}/N(t)$.
Next, $N^\ast(t+1)$ offspring individuals are sampled by sampling parental
pairs and gametes proportional to $w_{ij}^{kl}$, and sampling a trait value
in accordance with \cref{eq:oneline}.

@@

Individual-based simulations for the mixed-ploidy infinitesimal model were
implemented in Julia \citep{julia}.
Documented code and simulation notebooks are available at
\url{https://github.com/arzwa/InfGenetics}.

@@

While the probability of establishment becomes lower as the strength of
selection becomes large, the establishment probability does not decrease
monotonically with $\gamma$, i.e. as selection becomes very weak, the
establishment probability also decreases.
This happens because, although the probability of surviving the first couple of
generations becomes higher when selection is weaker, adaptation (i.e. increase
in the trait mean) will be slower,
increasing the risk that genetic variation is exhausted due to inbreeding before the
population is able to reach a consistently positive growth rate.
Note that when $\gamma = 0$, each individual has fitness 1 and we have a
critical branching process with a Poisson offspring distribution, so that the
probability to reach $N=100$ is $1/100$ \citep{barton2018}.

@@

Importantly, we assume no intrinsic advantage or disadvantage of polyploids in
the marginal habitat, i.e. the expected fitness of a migrant individual is the
same regardless of the ploidy level.
Differences in establishment probabilities are hence caused solely by aspects
of autopolyploid genetics and the barrier to gene flow between subpopulations
of different ploidy levels.
This is undoubtedly unrealistic.
For instance, neopolyploids are likely to suffer intrinsic fertility issues due
to meiotic irregularities associated with multivalent formation
\citep{bomblies2016,novikova2023}.
Furthermore, in our simulations, we have focused on the case where the
equilibrium genetic variance is constant across cytotypes ($\beta_4=\sqrt{1/2}$
in our model).
It would be very interesting to experimentally assess whether typical
quantitative traits in mixed-ploidy populations follow approximately the
mixed-ploidy infinitesimal model we outlined in this paper, and to actually
estimate how the variance scales across ploidy levels.
In addition, we have assumed a relatively high and constant rate of unreduced
gamete formation $u$ and triploid fertility $v$ in all our simulations (5\%),
whereas these are known to be variable across the population, and at least in
part genetically determined \citep{kreiner2017,clo2022c}.
We ignore such complications, and hence do not take the actual establishment
probabilities very serious, focusing instead on how migration load and
prezygotic isolation affect the tetraploid establishment probability.

@@

In the presence of maladaptive gene flow, a nascent tetraploid subpopulation
suffers from MCE, and although polyploids are more reproductively isolated from
a typical migrant (and hence suffer less maladaptive gene flow), MCE will
increasingly hamper the establishment of tetraploids as the rate of migration
grows.
Nevertheless, it is important to remark that despite MCE, the probability of
tetraploid establishment in the marginal habitat can be an order of magnitude
higher than expected based on the frequency of tetraploid migrants (i.e. is
roughly $O(u)$ instead of $O(u^2)$) when migration is sufficiently weak and
maladaptation sufficiently high.

@@

A major weakness of the present work, and an important caveat, is that we have
ignored inbreeding depression and dominance throughout.
Including dominance in the infinitesimal framework is already challenging for
diploids (requiring the tracking of four-way identity coefficients;
\cite{barton2023}), and appears intractable for higher ploidy levels.
However, autopolyploidy has important consequences whenever dominance is
relevant, as in the case of inbreeding depression \citep{ronfort1999}.
Indeed, when inbreeding depression is due to recessive deleterious variation,
it is expected to be less expressed in neotetraploids because homozygous genotypes
should be much rarer than in their diploid parents.
When there is indeed deleterious recessive variation, inbreeding during the
establishment process should therefore incur a higher fitness cost in diploids
relative to tetraploids, and hence further increase the probability of
tetraploid establishment.
How this plays out depends on the \textit{rate} of inbreeding, which will
differ between cytotypes. 
In outcrossing populations, inbreeding occurs at a slower rate in tetraploids,
further decreasing inbreeding depression and aiding tetraploid establishment.
However, when polyploidization is associated with increased selfing (as when it
disrupts an existing SI system), increased inbreeding depression in
autotetraploids may prevent their establishment.

@@

In the long term, polyploids are expected to accumulate a larger
mutation load when deleterious variation is recessive due to less efficient
purging, and this may yield \textit{increased} inbreeding depression
\citep{vlcek2025biorxiv}.
These effects have been studied in the context of range expansions
\citep{booker2024}.
However, this applies only to polyploids that have been established for a long
time.
In our case, polyploids are always recently descended from diploid ancestors,
and they will not have accumulated more deleterious mutations than their
diploid counterparts, so that polyploidy leads to reduced rather than increased
inbreeding depression when selfing rates are similar (as discussed above).
Interestingly, the interplay between the effects of polyploidy on different
timescales could yield an equilibrium situation that may characterize
mixed-ploidy populations in nature: although sometimes polyploids could enjoy
enhanced establishment probabilities in peripheral habitats, the accumulation
of mutational load may in the long-term limit further range expansion or
even lead to competitive exclusion by diploids.
Further modeling efforts could provide more insights into the plausibility of
such a model.

\section*{Acknowledgements}

I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
version.
I acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
Postdoctoral Fellowship 1272625N).

>>>>

\documentclass[12pt,a4paper]{article}
\input{preamble.tex}
\usepackage[margin=1in]{geometry}
\usepackage{authblk}
\usepackage{lineno}
\linenumbers
\externaldocument[s-]{supp}

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
%\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
%\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
\author{}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics for the inheritance
    of an additive quantitative trait in a mixed-ploidy population consisting
    of diploid, triploid and autotetraploid individuals producing haploid and
    diploid gametes.
    We implement efficient simulation methods and use these to study the
    quantitative genetics of mixed-ploidy populations and the establishment of
    autotetraploids in a new habitat.
    We show that for rare migration from the source population, autotetraploids
    are more likely to found a successful population than diploids under a very
    broad range of conditions, but that this is unlikely to sufficiently
    counter the scarcity of tetraploid founders when the source is
    predominantly diploid.
    We assess in more detail how minority cytotype exclusion interacts with
    migration load in the establishment process and evaluate the impact of
    additional sources of prezygotic isolation, specifically selfing and
    assortative mating, on the relative establishment probabilities of the
    different cytotypes.
    In the discussion we consider how inbreeding depression may impact our
    findings.

@@

@@

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population through local extinction and 
recolonization dynamics has been suggested as a plausible model
\citep{lande1985}.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for local tetraploid establishment to
occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of \textit{prezygotic isolation} between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, 
localized dispersal \citep{baack2005,kolar2017} and asexual reproduction
\citep{vandrunen2022}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression
\citep{ronfort1999,otto2000,husband2008,clo2022b}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

@@

Our notation is summarized in \cref{tbl:glossary}.
We consider a mixed-ploidy population of size $N$ consisting of $N_2$ diploid,
$N_3$ triploid and $N_4=N-N_2-N_3$ tetraploid individuals.
We assume an individual of ploidy level $k$ forms haploid and diploid gametes
with proportions $u_{k1}$ and $u_{k2}$, as well as a proportion
$1-u_{k1}-u_{k2}$ inviable (e.g. aneuploid or polyploid) gametes.
The (relative) fecundity of a $k$-ploid individual is hence $u_{k1} + u_{k2}$.
Unless stated otherwise, we will assume 
\begin{equation}
    \begin{pmatrix} 
    u_{21} & u_{22} \\ 
    u_{31} & u_{32} \\ 
    u_{41} & u_{42} 
    \end{pmatrix} =
    \begin{pmatrix} 
    1-u & u \\
    v & v \\
    0 & 1-u
    \end{pmatrix} \label{eq:U}
\end{equation}
where $u$ is referred to as the proportion of unreduced gametes, and $2v$ is
the proportion of euploid gametes produced by a triploid individual.
The rate of unreduced gamete production is hence fixed across individuals, and
is assumed to be the same in diploids and tetraploids.

@@

If we would naively assume that the allelic effects underlying an additive
trait are identical across ploidy levels, 
a tetraploid offspring from a cross between two diploids would have, on average,
a trait value which is the sum of the parental trait values.
This is not likely to reflect biological reality: tetraploids do not tend to
have, for instance, twice the size of their diploid progenitors on average
(e.g. \cite{porturas2019}).
Furthermore, the genetic variance at Hardy-Weinberg and linkage equilibrium
(HWLE) in a large non-inbred tetraploid population will be twice that of their
diploid counterparts under such assumptions, which is similarly unrealistic
\citep{clo2022}.

@@

Again following \cite{barton2018}, we simulate the model by first calculating
the fitness of each parental pair $(i,j)$, which is the expected fitness of
offspring of this pair
\begin{equation}
  w_{ij}
    = \sum_{k=1}^2\sum_{l=1}^2 w_{ij}^{kl}
    = \sum_{k=1}^2\sum_{l=1}^2 u_{c_{i},k}u_{c_{j},l}
        \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right]
\end{equation}
The expectation on the right hand side can be calculated from \cref{eq:oneline}
using the moment-generating function of the Gaussian, i.e.
\begin{equation}
    \Ex\left[e^{\gamma(Z_{ij}^{kl} - \theta)}\right] 
    = e^{\gamma \left(\Ex \left[Z_{ij}^{kl}\right] -\theta\right) +
    \frac{\gamma^2}{2}
    \Var\left[Z_{ij}^{kl}\right]} 
\end{equation}
Having calculated the $w_{ij}$, the number of offspring surviving into the next
generation is calculated as $N^\ast(t+1) = \sum_{i,j}w_{ij}/N(t)$.
Next, $N^\ast(t+1)$ offspring individuals are sampled by sampling parental
pairs and gametes proportional to $w_{ij}^{kl}$, and sampling a trait value
in accordance with \cref{eq:oneline}.

@@

Individual-based simulations for the mixed-ploidy infinitesimal model were
implemented in Julia \citep{julia}.
Documented code and simulation notebooks are available at \url{[anonymized]}.
%\url{https://github.com/arzwa/InfGenetics}.

@@

While the probability of establishment becomes lower as the strength of
selection becomes large, the establishment probability does not decrease
monotonically with $\gamma$, i.e. as selection becomes very weak, the
establishment probability also decreases.
This happens because, although the probability of surviving the first couple of
generations becomes higher when selection is weaker, adaptation (i.e. increase
in the trait mean) will be slower,
increasing the risk that genetic variation is exhausted due to inbreeding before the
population is able to reach a consistently positive growth rate.
Note that when $\gamma = 0$, each individual has fitness 1 and we have a
critical branching process with a Poisson offspring distribution, so that the
probability to reach $N=100$ is $\sim 1/100$ \citep{barton2018}.

@@

Importantly, we assume no intrinsic advantage or disadvantage of polyploids in
the marginal habitat, i.e. the expected fitness of a migrant individual is the
same regardless of the ploidy level.
Differences in establishment probabilities are hence caused solely by aspects
of autopolyploid genetics and the barrier to gene flow between subpopulations
of different ploidy levels.
This is undoubtedly unrealistic.
For instance, neopolyploids are likely to suffer intrinsic fertility issues due
to meiotic irregularities associated with multivalent formation
\citep{bomblies2016,novikova2023}, and triploids may be inviable due to
issues with endosperm development \citep{bretagnolle1995}.
Similarly implausible is the assumption of a constant equilibrium genetic
variance across cytotypes ($\beta_4=\sqrt{1/2}$ in our model), which we used in
most of our results.
It would be very interesting to experimentally assess whether typical
quantitative traits in mixed-ploidy populations follow approximately the
mixed-ploidy infinitesimal model we outlined in this paper, and to actually
estimate how the variance scales across ploidy levels.
In addition, we have assumed a relatively high and constant rate of unreduced
gamete formation $u$ and triploid fertility $v$ in all our simulations (5\%),
whereas these are known to be variable across the population, and at least in
part genetically determined \citep{kreiner2017,clo2022c}.
We ignore such complications, and hence do not take the actual establishment
probabilities very serious, focusing instead on how migration load and
prezygotic isolation affect the tetraploid establishment probability.

@@

In the presence of maladaptive gene flow, a nascent tetraploid subpopulation
suffers from MCE, and although polyploids are more reproductively isolated from
a typical migrant (and hence suffer less maladaptive gene flow), MCE will
increasingly hamper the establishment of tetraploids as the rate of migration
grows.
Nevertheless, it is important to remark that despite MCE, the probability of
tetraploid establishment in the marginal habitat can be an order of magnitude
higher than expected based on the frequency of tetraploid migrants (i.e. is
roughly of order $u$ instead of $u^2$) when migration is sufficiently weak and
maladaptation sufficiently high.

@@

A major weakness of the present work, and an important caveat, is that we have
ignored inbreeding depression and dominance throughout.
Including dominance in the infinitesimal framework is already challenging for
diploids (requiring the tracking of four-way identity coefficients;
\cite{barton2023}), and appears intractable for higher ploidy levels.
However, autopolyploidy has important consequences whenever dominance is
relevant, as in the case of inbreeding depression
\citep{ronfort1999,husband2008,clo2022b}.
Indeed, when inbreeding depression is due to recessive deleterious variation,
it is expected to be less expressed in neotetraploids because homozygous genotypes
should be much rarer than in their diploid parents (the `masking' effect;
\cite{otto2000}).
In the presence of deleterious recessive variation, inbreeding during the
establishment process should therefore incur a higher fitness cost in diploids
relative to tetraploids, and hence further increase the probability of
tetraploid establishment.
How this plays out depends however on the \textit{rate} at which populations
become inbred, which will differ between cytotypes and will depend strongly on
the mating system.
In outcrossing populations, inbreeding occurs at a slower rate in tetraploids,
further decreasing inbreeding depression and aiding tetraploid establishment.
However, when polyploidization is associated with increased selfing (as when it
disrupts an existing SI system), increased inbreeding depression in
autotetraploids may prevent their establishment.

@@

In the long term, polyploids are expected to accumulate a larger
mutation load when deleterious variation is recessive due to less efficient
purging, and this may yield \textit{increased} inbreeding depression
\citep{vlcek2025biorxiv}.
These effects have been studied in the context of range expansions
\citep{booker2024}.
However, this applies only to polyploids that have been established for a long
time.
In our case, polyploids are always recently descended from diploid ancestors,
and they will not have accumulated more deleterious mutations than their
diploid counterparts, so that polyploidy should lead to reduced rather than
increased inbreeding depression when selfing rates are similar (as discussed
above).
Interestingly, the interplay between the effects of polyploidy on different
timescales could yield an equilibrium situation that may characterize many
mixed-ploidy populations in nature: although sometimes polyploids could enjoy
enhanced establishment probabilities in peripheral habitats, the accumulation
of mutational load may in the long-term limit further range expansion or
even lead to competitive exclusion by diploids.
Further modeling efforts could provide more insights into the plausibility of
such a model.

%\section*{Acknowledgements}
%
%I thank Christelle Fraïsse for encouragement and feedback. Thanks to Quinten
%Bafort, Felipe Kauai and Frederik Mortier for comments on an early draft
%version.
%I acknowledge funding from the Research Foundation -- Flanders (FWO, Junior
%Postdoctoral Fellowship 1272625N).
